---
title: "proteins"
author: "Gerardo José Rodríguez"
date: "2022-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Biostrings)
library(dplyr)
library(ggplot2)
library(tidyverse)
```
---
title: "proteins"
author: "Gerardo José Rodríguez"
date: "2022-10-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Biostrings)
library(dplyr)
library(ggplot2)
library(tidyverse)
```

## R Markdown
Lets consider a native sequence and make a mutagenesis library. For that we will need a random-substitution function.

```{r}
# Define the initial sequence
nat =  DNAString("ATGGGTAACATCTAA")

# Alternatively read it from a FASTA file
#nat = readDNAStringSet('native_DNA.fa')[[1]]

# Define a substitution function
make_substitution = function(seq, site=FALSE, nucl=FALSE){
  if (site==FALSE){site=sample(c(1:nchar(seq)),1)}
  if (nucl==FALSE){nucl=sample(c('A','C','G','T'),1)}
  c(seq=replaceLetterAt(seq, at=site, nucl),
    mut=c(paste0(site, nat[site],'>',nucl)))}

# Perform the function over a set of your original sequence
set = rep(DNAStringSet(nat),1000)
set = lapply(set, \(dna) make_substitution(dna)$seq)
```

### Determine frequency at each position of each nucleotide

```{r}
cm_dna = consensusMatrix(DNAStringSet(set))[DNA_BASES,]
cm_dna
```


### Determine frequency at each position of each residue
```{r}
# Translate your DNA set to a protein set
prot_set = AAStringSet(lapply(DNAStringSet(set), \(x) translate(x)))

resFreqPerPos = function(prot_set){
  # Build a count matrix
  cm_res = consensusMatrix(AAStringSet(prot_set))
  # Determine which AA are not represented to complete the table content with 0s
  AAtoAdd = setdiff(AA_ALPHABET, rownames(cm_res))
  rbind(cm_res, matrix(0, nrow=length(AAtoAdd), ncol=ncol(cm_res),
                       dimnames=list(AAtoAdd, 1:ncol(cm_res))))
}
resFreqPerPos(prot_set)
```

# Determine each residue-pair substitution frequence
```{r}
# I CREATED TWO POTENTIAL FUNCTIONS. THE SECOND ONE IS ALMOST 5x FASTER, BUT DO NOT CREATE STANDARD TABLES (ONLY NECESSARY AMINOACIDS WITH AT LEAST ONE COUNT)
nat_prot = translate(nat)

subMtx = function(prot_set, nat_prot){
  # Create a count pair-matrix
  sub_mtx = matrix(0, nrow=length(AA_ALPHABET), ncol=length(AA_ALPHABET),
                   dimnames=list(AA_ALPHABET, AA_ALPHABET))
  # Find the residue-pairs in the substitutions and summarize them in a table
  pairs = table(unlist(sapply(prot_set, function(prot){
    pos = which(as.integer(nat_prot)!=as.integer(prot), useNames = T)
    paste0(as.vector(nat_prot[pos]), as.vector(prot[pos]))
  })))
  # Fill the matrix values with the correspondant counts for each pair
  for (p in 1:length(pairs)){
    sub_mtx[substr(names(pairs)[p],1,1), substr(names(pairs)[p],2,2)] = pairs[p]
  }
  sub_mtx
}

subMtx2 = function(prot_set, nat_prot=NULL){
  
  # Create a consensus protein (in case is not given)
  if (is.null(nat_prot)){nat_prot=consensusString(prot_set)}
  
  # Align sequences and find all mismatches. Reshape the object in a mtx
  prot_set_aligned = pairwiseAlignment(prot_set, nat_prot) 
  sub_mtx = mismatchSummary(prot_set_aligned)$subject %>%
    group_by(Subject, Pattern) %>%
    summarize(Count=sum(Count)) %>%
    xtabs(data=., formula=Count~Subject+Pattern)
    AAtoAdd_row = setdiff(AA_ALPHABET, rownames(sub_mtx))
    AAtoAdd_col = setdiff(AA_ALPHABET, colnames(sub_mtx))
    sub_mtx[AAtoAdd_row,] = 0
    sub_mtx[AAtoAdd_row,] = 0
    rbind(sub_mtx, matrix(0, nrow=length(AAtoAdd), ncol=ncol(sub_mtx),
                       dimnames=list(AAtoAdd, 1:ncol(sub_mtx))))
  }

dim(subMtx(prot_set, nat_prot))
dim(subMtx2(prot_set, nat_prot))
```

## Make some relevant plots of the results
```{r}
library(pheatmap)
# Heatmap of pair-substitution frequence
heatmap(subMtx2(prot_set, nat_prot)[AA_STANDARD, AA_STANDARD], 
        Rowv = NA, Colv = NA)

pheatmap(subMtx2(prot_set, nat_prot)[AA_STANDARD, AA_STANDARD])

# Heatmap of residue presence per position
heatmap(resFreqPerPos(prot_set),
        Rowv = NA, Colv = NA)
```



Now we would like to extract some information from these sequences. Let's see if we can create a function identyfying the mismatches between two sequences

```{r}
# Function where you introduce 
find_subs_dna = function(dna1, dna2, do_align=FALSE){
  positions = which(!as.integer(dna1)==as.integer(dna2))
  sapply(positions, \(pos) paste0(pos, dna1[pos], '>', dna2[pos]))
}
# Apply through all the sequences
unlist(sapply(set, \(dna) find_subs_dna(nat, dna)))
```



# ALL FUNCTIONS
## In case you want to use them without analyzing them one by one
```{r}
make_substitution = function(seq, site=FALSE, nucl=FALSE){
  if (site==FALSE){site=sample(c(1:nchar(seq)),1)}
  if (nucl==FALSE){nucl=sample(c('A','C','G','T'),1)}
  c(seq=replaceLetterAt(seq, at=site, nucl),
    mut=c(paste0(site, nat[site],'>',nucl)))}

freq_mtx = function(setSeqs, natProt=NULL, justSubs=F, as.prob=T){
  # Determine if there is a DNA or protein set
  if(class(setSeqs)=='AAStringSet'){alphabet=AA_ALPHABET}
  else if(class(setSeqs)=='DNAStringSet'){alphabet=DNA_ALPHABET}
  else{return(print('Introduce a Biostrings AAStringSet/DNAStringSet'))}
  
  # Determine residue's frequence at each position
  freq_mtx = as.matrix(setSeqs) %>% 
  as.data.frame() %>%
  apply(., 2, function(col){
  table(c(col, alphabet))-1}) 
  
  # Name columns as sequence-position + letter (if native seq is specified)
  if (is.null(natProt)){colnames(freq_mtx) = 1:ncol(freq_mtx)}
  else {colnames(freq_mtx) = paste0(1:nchar(natProt), as.vector(natProt))}
  
  # Keep or not the equal subs
  if (!is.null(natProt) && justSubs==T){
    for (l in 1:nchar(natProt)){freq_mtx[as.character(as.vector(natProt)[l]),l] =0 }
  }
  # Return the output
  if (as.prob==T) {return(freq_mtx/sum(freq_mtx))}
  else{return(freq_mtx)}
}  

sub_mtx = function(setSeqs, wildSeq, justSubs=F){
    # Determine if there is a DNA or protein set
  if(class(setSeqs)=='AAStringSet'){alphabet=AA_ALPHABET}
  else if(class(setSeqs)=='DNAStringSet'){alphabet=DNA_ALPHABET}
  else{return(print('Introduce a Biostrings AAStringSet/DNAStringSet'))}
  
  # Determine residue's frequence at each position
  freq_mtx = as.matrix(setSeqs) %>% 
  as.data.frame() %>%
  apply(., 2, function(col){
  table(c(col, alphabet))-1}) 

  # Group columns that share the same letter in the native sequence
  colnames(freq_mtx) = as.vector(wildSeq)
  sub_mtx = t(rowsum(t(freq_mtx), group=colnames(freq_mtx)))  # Returns sub_mtx
  
  # If just substitutions are of interest discard the concordances
  if (justSubs){for (l in colnames(sub_mtx)){sub_mtx[l,l] = 0 }}
  
  # Add lacking aminoaciads in columns
  LettersToAdd = setdiff(rownames(sub_mtx), colnames(sub_mtx))
  cbind(sub_mtx, matrix(0, nrow=nrow(sub_mtx), ncol=length(LettersToAdd),
                       dimnames=list(rownames(sub_mtx),
                                     LettersToAdd)))
}  
```

